Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-02 versions/response.v1.txt

> The other main concern that I have is with the restriction that
> t=(x,y,x,y,w) in estimation rather than t=(t_1,...,t_5). On the one
> hand, usually one expects that incorporating knowledge about the true
> underlying tree (t_1=t_3 & t_2=t_4 in the present case) should be
> helpful for a good estimation procedure. Thus the fact that treating
> ancestral states as parameters gives inconsistent topological
> estimation in this setting demonstrates that the approach is
> problematic. On the other hand, allowing t=(t_1,...,t_5) is what would
> be done in any practical implementation and it seems quite possible to
> me that results would be radically different in this case. For
> instance, from Fig 2, the generating tree with all equal edge-lengths
> (y*=x*=0.2) provides a case where topology tau_2 is inconsistently
> estimated with the restriction that estimated t=(x,y,x,y,w). Because
> of the symmetry of the problem, I am doubtful y*=x*=0.2 would be a
> point of inconsistency if t=(t_1,...,t_5) were unrestricted in
> estimation. The issue deserves discussion because I don't think it is
> transparent.

TODO: make the argument more clear in the main text.

> I don't think the restricted case of p8-9 will be of interest to
> readers. Only the general case.

%EM I think we could move this to the appendix to throw this reviewer a bone.
%das: does this reviewer mean both the general case and the empirical case or just the empirical case?
%Their review makes me think they take the empirical case the be the "general" case.
%You think we should move both analytic cases to the appendix or just the one?

@@

> Also, laying out the theory a little bit more step-by-step so that a
> general quantitative reader doesn't struggle too much to follow would
> help (suggestions below).

Thank you very much for reading the derivations and for your suggestions. As noted inline we have revised the text accordingly.

> I know that it's not feasible to explicitly compute (analytically)
> these likelihoods for bigger trees. But one of the main motivations
> and applications for tools like TreeTime is surely that the number of
> sequences grows (not the number of sites, which is fixed by the viral
> sequence length in the case of viruses). What happens when the number
> of tips grows? Are there more opportunities for the kind of issue that
> gives the wrong value for w? Many influenza trees have many 0 (or near
> 0) internal branch lengths, and it would be interesting to understand
> the implications for the ancestral state reconstructions (which in
> turn are used to infer spatial spread, transmission etc).

%EM David-- does it seem like we could extend these results to bigger trees containing a 4 taxon tree?
%das: I don't see why not, though I haven't thought it through very much.
%It could take a lot of work or not much work at all.
%Do you know of any other mathematical phylogenetics papers that do something like this?
%Like induction on trees?
%I can think a bit and get back to you...

@@

> Some detailed comments and questions:
> – line 85 some cut/paste or sentence issue

TODO: not sure what this is?
%das: this reviewer was confused about the phrase "generating probabilities," thinking we were generating the probabilities and not that they were the probabilities of generating a particular data configuration.
%I changed:
%Fidelities have useful algebraic properties, and generating probabilities using the Hadamard transform have an especially simple form (see \eqref{eq:hadamard_probability} in the Appendix).
%to:
%Fidelities have useful algebraic properties.
%As data becomes plentiful, we use the Hadamard transform (see \eqref{eq:hadamard_probability} in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form.

We have added the definition of a "generating probability" to ease confusion in this sentence.

@@

> L354 why does psi take the complement of site patterns? Just above it
> is defined as a random variable psi (Y) where Y is a site pattern

This wasn't as clear as it should have been. We have revised to:
"The mapping $\patternToSplit$ operates by returning the complement of a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if it is not in $\mathcal{P}(\siteSplitSet)$."

> What are the elements of the sets eta(tau, t)?

TODO: fill in
%das: This reviewer was confused about the bold eta, which was an ordered list of sets eta_1, eta_2, ...
%I noticed we never actually used this ordered list so got rid of it and wrote:
%Thus given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies...

We removed the ordered list of sets as it did not add to any understanding of the concept and now refer to each of these sets and their relevance directly.

@@

> line 364 define p(e)

Thanks for reminding us to define p(e), which is now described as "where $p(e)$ is the probability of a character change along edge $e$."

> line 366  I don't understand (8) but perhaps that's just background I
> don't have – a quick explanation or reference would be handy
> What's a 'generating probability' ? Is that the fidelity? Or a
> generator?

TODO: just say tough nuts?
%das: I defined "generating probability" earlier, but, I mean, come on, a reference? We give a section of a chapter! I'd vote for tough nuts:
%We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.

We have updated the text to explicitly define "generating probability." Background for equation (8) can be found in Section 8.6 in the Semple and Steel text. This is referenced at the beginning of the section.

@@

> Reviewer: 3

> Comments to the Author
> Shaw and Matsen show that a joint maximum-likelihood inference of
> phylogenies and ancestral states is not consistent when branch lengths
> are long. We have recently proposed such a joint estimation for
> sequences that recently diverged and are hence similar (Sagulenko et
> al). The results presented here are anticipated in Sagulenko et al
> "For more divergent sequences, however, subleading states of internal
> nodes make a substantial contribution and the iterative optimization
> will underestimate the branch lengths."
> and
> "The iterative optimization procedures are not expected to be accurate
> for trees were many sites are saturated. In scenarios with extensive
> uncertainty of ancestral states and tree topology, convergence of the
> iterative steps cannot be guaranteed."

> In this light, it should be pointed out that joint estimation
> procedure was explicitly suggested as an approximation and hence never
> expected to be consistent in a strict sense. Furthermore, it was
> suggested as an approximation when branch lengths are short -- the
> opposite of the problematic regions identified here.
> But systematic analysis of the behavior of this approximation is
> useful and I believe the work is correct (but haven't checked the
> proofs). The work by Shaw and Matsen suggests the radius of
> convergence of this procedure is somewhere around branch length of
> order 10-20%, which is consistent with results reported by Sagulenko
> et al.

Thank you for your thoughts, and we have worked hard to not oversell our results.
TODO we should make some changes to at least say that we've further clarified the implications of our results.
%das: can you clarify this TODO?
%Do you mean we should add more text about the bias in Fig. S3, or we should quote Richard more and show how we think things might still be bad?
The TreeTime paper does present joint optimization as an approximate approach, and described that it is for short branch lengths.
However, in our assessment the word "approximate" doesn't completely describe a situation in which there is significant bias.
The TreeTime paper says that its use is appropriate when "ancestral sequence inference has little ambiguity" but even in these cases the bias in branch lengths persists.

@@

Thank you for your assessment.
We have worked hard in our initial submission and in our revision to write out the argument in great (indeed, excruciating) detail, and we would appreciate it if you could take a look at the revision.

> The only real criticism I have of the paper is that the authors could
> have done a better job in connecting their new work to previous
> results.  For example, Goldman (1990, Syst. Biol. 39:345-361) showed
> that joint estimation of topology and ancestral states by maximum
> likelihood was equivalent to maximum parsimony analysis for the same
> 2-state model considered by Shaw and Matsen when all branches are
> forced to have the same length.  They noted the statistical
> inconsistency of this approach, which was necessarily true because
> parsimony was already known to be inconsistent (the famous result of
> Felsenstein 1978).  Shaw and Matsen extend this result to the case of
> unequal branch lengths, which while not particularly surprising, is
> nonetheless an important extension.  The authors should also cite key
> relevant earlier work on the identifiability ot the tree topology with
> "classical" maximum likelihood phylogenetic estimation (Allman et al
> 2008, Adv Appl Prob 40:229–249; Chai and Housworth 2011, Syst Biol
> 60:713–718) in addition to crediting Roychoudhury 2015 for
> establishing ML consistency.

We are very grateful to this reviewer for pointing out the Goldman paper.
We were not aware of this paper and now have a paragraph describing it and comparing to our work.
It's also a good idea to cite additional papers about identifiability, and we have done so.

%EM David, heh, can you point me to the place where the equal branch length result was done?
%EM Also, I think you said that he says that the varying branch length thing is an open question-- it would be great to quote that.
%das: below are some notes and quotes. I will work on paring these down and adding some to the text, maybe with a little more diplomacy.
%The equal branch length example occurs in Example 2, starting halfway down the second column of page 352 in Goldman (1990) and continuing until the top of the second column on page 354.
%Here he only talks about how parsimony is inconsistent because it may result in equally parsimonious trees (this is the last paragraph of the example section).
%It's not until the bottom of page 356 that he punts the inconsistency to Felsenstein and Hendy/Penny, and the bottom of page 357 where he wonders how small branch lengths have to be for parsimony to be consistent in general.

---

Goldman (1990) concerns estimating both structural (e.g., topology, branch lengths) and incidental (e.g., ancestral states) parameters simultaneously, though the only worked example is in the case of a topology with equal branch lengths.
Goldman shows that for certain special cases, "maximum relative likelihood," i.e., performing maximum likelihood over both topology and ancestral states, is equivalent to parsimony, and thus inconsistent.
He says
"Consistent estimates of incidental parameters are unobtainable, as it is not possible to increase indefinitely the number of observations (data) pertinent to each."
However, he also says "results may be unpredictable" and "estimates of the structural parameters (e.g., the tree T) may also be inconsistent."
He does not discuss the inconsistency of maximum relative likelihood in general, and proceeds to show that parsimony analysis is a specific instance of maximum relative likelihood, implying maximum relative likelihood of topology given equal branch lengths cannot be consistent in general, but stops there.
We show cases where the estimates of the topology will be inconsistent if we are estimating both topology and branch lengths, and, moreover, even if we know the correct topology, we show cases where branch lengths will be inconsistent.
Finally, just prior to his conclusion, he discusses when parsimony will give the same answer as ML, concluding that the question is ill-posed since parsimony estimates different parameters than ML, i.e., it assumes equal branch lengths.
He states that if the expected amount of change along a branch is "very small" then parsimony will recover the same tree as ML (under a Poisson process model).
We are able to quantify how small is "very small" in the case of general joint ML---estimating topology and branch lengths together---and not just parsimony in particular.

quote:
"The second approach is to estimate the random variables as though they were parameters of the model. However, in this case they will be incidental parameters: as the amount of data (i.e., the number of characters) increases, the number of parameters also increases. For each additional character (labeled a, for instance), there are additional data xa, consisting of the states for all n species, and additional parameters Ya, consisting of the states for all the internal nodes {NI: I = 1,2, . . , m}. Although this method is the only one possible if we wish to estimate the internal node character states, the results may be unpredictable. As noted above, these estimates cannot be consistent. Further, the estimates of the structural parameters (e.g., the tree T) may also be inconsistent. Indeed, under certain models, this approach renders inferences virtually impossible (Cavalli-Sforza et al., 1964; Cavalli-Sforza and Edwards, 1966; Kendall and Stuart, 1979). Examples of both approaches are investigated below."

---

### 2018-05-02 appendix.tex
everybody hates farris

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
    $\{1\}$     & $p_{1}$   &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{2\}$     & $p_{2}$   &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    $\{3\}$     & $p_{3}$   &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{1,2\}$   & $p_{12}$   &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,3\}$   & $p_{13}$   &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
    $\{2,3\}$   & $p_{23}$   &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,2,3\}$ & $p_{123}$   &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\subsection*{Example}
We follow with an expository example computing these probabilities and likelihoods.
Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a---this is commonly known as the ``Farris zone'' topology.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
\[
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
\]
Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
\]
in general.

@@

To compute the likelihood of observing a set of data, we need $\Pr(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
Using branch fidelities, the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the Farris tree.
Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} (for the Farris tree) and~\ref{tab:fels_likelihoods} (for the Felsenstein tree).
Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}.

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_like00}
\caption[short]{}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_like10}
\caption[short]{}
\end{subfigure}
\caption{
    Example likelihood computations on the Farris tree $\tau_1$ for fidelities $x$, $y$, and $w$.
    Edges labeled by the probability of substitution along that edge.
    In (a), we compute the product to obtain $\Pr(\ancestralStateColumn=\emptyset\mid \alignmentColumn=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1+w)/32$.
    In (b), the same process yields $\Pr(\ancestralStateColumn=\{1\}\mid \alignmentColumn=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1-w)/32$.
}
\label{fig:example_likelihoods}
\end{figure}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$\Pr(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
    \hline
     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$          &$(1+x)^2   (1-w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1-x)^2   (1+w)(1+y)^2$          &$(1-x)^2   (1-w)(1-y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1-x)^2   (1+w)(1+y)(1-y)$       &$(1-x)^2   (1-w)(1+y)(1-y)$\\
    \hline
    \hline
    &$\{1\}$                             &$\{1,2\}$  \\
    \hline
     $\emptyset$   &$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1+x)^2   (1-w)(1+y)^2$     &$(1+x)^2   (1+w)(1-y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1+x)^2   (1-w)(1+y)(1-y)$  &$(1+x)^2   (1+w)(1+y)(1-y)$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the Farris tree $\tau_1$.
All values multiplied by $1/32$.}
\label{tab:farris_likelihoods}
\end{table}

@@

    \multicolumn{3}{c}{Farris tree ($\tau=\tau_1$)}\\

@@

\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and ancestral state partitions $\ancestralSplitPartition_j$ after maximizing over ancestral states on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$.

@@

Assume $\tau^*=\tau_1$.
The Farris tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see the upper table of Table~\ref{tab:likelihoods}).

@@

We take various safeguards to ensure convergence and stability of the procedure.
Since this method involves randomness, we take precautions to not evaluate the likelihood near or outside of boundary conditions, i.e., when $x^*$ or $y^*$ are either $0$ or $1$.
As is apparent from Table~\ref{tab:sitepatprob}, if $0 < x^*, y^* < 1$ and we try to evaluate the candidate $\hat{x}=\hat{y}=\hat{w}=1$ we have a likelihood of exactly zero.
Worse still, in Table~\ref{tab:likelihoods} for the Farris tree, in the cases of $\{2\}, \{1,2\}, \{2,3\},$ and $\{1,2,3\}$, any attempt to evaluate $\hat{y}=1$ will similarly yield a zero likelihood.
In these cases of zero likelihood, methods that evaluate many candidate parameter values can fail to converge when near these boundaries.
We sidestep these computational issues by restricting ourselves to the region $x^*,y^*\in [10^{-2},1-10^{-2}]^2$.

>>>>

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
    $\{1\}$     & $p_{1}$   &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{2\}$     & $p_{2}$   &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    $\{3\}$     & $p_{3}$   &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{1,2\}$   & $p_{12}$   &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,3\}$   & $p_{13}$   &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
    $\{2,3\}$   & $p_{23}$   &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,2,3\}$ & $p_{123}$   &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\subsection*{Example}
We follow with an expository example computing these probabilities and likelihoods.
Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
\[
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
\]
Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
\]
in general.

@@

To compute the likelihood of observing a set of data, we need $\Pr(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
Using branch fidelities, the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the InvFels tree.
Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} (for the InvFels tree) and~\ref{tab:fels_likelihoods} (for the Felsenstein tree).
Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}.

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_like00}
\caption[short]{}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_like10}
\caption[short]{}
\end{subfigure}
\caption{
    Example likelihood computations on the InvFels tree $\tau_1$ for fidelities $x$, $y$, and $w$.
    Edges labeled by the probability of substitution along that edge.
    In (a), we compute the product to obtain $\Pr(\ancestralStateColumn=\emptyset\mid \alignmentColumn=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1+w)/32$.
    In (b), the same process yields $\Pr(\ancestralStateColumn=\{1\}\mid \alignmentColumn=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1-w)/32$.
}
\label{fig:example_likelihoods}
\end{figure}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$\Pr(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
    \hline
     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$          &$(1+x)^2   (1-w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1-x)^2   (1+w)(1+y)^2$          &$(1-x)^2   (1-w)(1-y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1-x)^2   (1+w)(1+y)(1-y)$       &$(1-x)^2   (1-w)(1+y)(1-y)$\\
    \hline
    \hline
    &$\{1\}$                             &$\{1,2\}$  \\
    \hline
     $\emptyset$   &$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1+x)^2   (1-w)(1+y)^2$     &$(1+x)^2   (1+w)(1-y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1+x)^2   (1-w)(1+y)(1-y)$  &$(1+x)^2   (1+w)(1+y)(1-y)$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
All values multiplied by $1/32$.}
\label{tab:farris_likelihoods}
\end{table}

@@

    \multicolumn{3}{c}{InvFels tree ($\tau=\tau_1$)}\\

@@

\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and ancestral state partitions $\ancestralSplitPartition_j$ after maximizing over ancestral states on the InvFels tree $\tau_1$ and the Felsenstein tree $\tau_2$.

@@

Assume $\tau^*=\tau_1$.
The InvFels tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see the upper table of Table~\ref{tab:likelihoods}).

@@

We take various safeguards to ensure convergence and stability of the procedure.
Since this method involves randomness, we take precautions to not evaluate the likelihood near or outside of boundary conditions, i.e., when $x^*$ or $y^*$ are either $0$ or $1$.
As is apparent from Table~\ref{tab:sitepatprob}, if $0 < x^*, y^* < 1$ and we try to evaluate the candidate $\hat{x}=\hat{y}=\hat{w}=1$ we have a likelihood of exactly zero.
Worse still, in Table~\ref{tab:likelihoods} for the InvFels tree, in the cases of $\{2\}, \{1,2\}, \{2,3\},$ and $\{1,2,3\}$, any attempt to evaluate $\hat{y}=1$ will similarly yield a zero likelihood.
In these cases of zero likelihood, methods that evaluate many candidate parameter values can fail to converge when near these boundaries.
We sidestep these computational issues by restricting ourselves to the region $x^*,y^*\in [10^{-2},1-10^{-2}]^2$.
